Tutorial: Fitting log-Gaussian mixed-effects mixture models for keystroke data

### 2024-05-19 tutorial/tutorial.Rmd
remove noncentreing

The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$). In the model below, these are estimated each level of a categorical predictor `location` for the $i^\text{th}$ interkey intervals.

@@

```{r mixturemodel, echo = F, fig.cap="What's a mixture model for interkey intervals? The parameters of the mixture model explained visually. Each of the three parameters can determined for each transition location."}

@@

Participants and conditions are expected to represent indices in the Stan code. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before-sentence = 1, before-word = 2, within-word = 3).  The returned values are indices for the parameters in the model: For example, for the location identifier, `beta[1]` is the population estimate for non-hesitant typing intervals at before-sentence transition locations, `beta[2]` for before-word transition locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.

@@

```{r }
posterior_ppts_long_diff_summary %>%
  mutate(across(difference, recode, 
                diff_sent = "Comparison: before-sentence vs before-word",
                diff_word = "Comparison: before-word vs within-word"),
         across(participant, as.numeric)) %>% 
  ggplot(aes(x = participant, 
             y = mean, 
             ymin = lower, 
             ymax = upper)) +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  geom_pointrange() +
  facet_wrap(~difference) +
  coord_flip() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Participant id", 
       y = "Difference in hesitation probabilities with 95% PIs") 
```

>>>>

The mixture model captures the writing process as a combination of three parameters: the average typing speed of fluent interkey intervals ($\beta$), the slowdown for hesitant interkey intervals ($\delta$) and the probability of hesitant interkey intervals ($\theta$). In the model below, the latter two are estimated separately for each level of the categorical prediction `location` (i.e. transition location) for the $i^\text{th}$ observation.

@@

```{r mixturemodel, echo = F, fig.cap="What's a mixture model for interkey intervals? The parameters of the mixture model explained visually."}

@@

Participants and conditions are expected to represent indices in the Stan code. In the code below we use `factor` in combination with `as.numeric` to ensure that there are no empty indices for participants and conditions. We create vectors with participant identifiers `ppts` (numbers $1$ through $I$ where $I$ is the number of participants) and a numeric identifier `condition` for each level of the transition location (levels: before sentence = 1, before word = 2, within word = 3). The returned values are indices for the parameters in the model: For example, for the location identifier, `delta[1]` is the population estimate for the slowdown for hesitant interkey intervals at before sentence transition locations, `delta[2]` for before word locations; `theta[1]` and `theta[2]` are the hesitation probability (on the logit scale) for before-sentence and before-word transition locations, respectively.

@@

```{r }
posterior_ppts_long_diff_summary %>%
  mutate(across(difference, recode, 
                diff_sent = "Comparison: before-sentence vs before-word",
                diff_word = "Comparison: before-word vs within-word"),
         across(participant, as.numeric)) %>% 
  ggplot(aes(x = participant, 
             y = mean, 
             ymin = lower, 
             ymax = upper)) +
  geom_hline(yintercept = 0, colour = "black", linetype = "dotted") +
  geom_pointrange() +
  facet_wrap(~difference) +
  coord_flip() +
  scale_x_continuous(breaks = 1:10) +
  labs(x = "Participant id", 
       y = "Difference in hesitation probabilities with 95% PIs") 
```
